F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-25 icfp-2015/paper.tex
First half of implementation

\subsection{Infering types from values}

@@

Then for all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, let $e_s=e[x\leftarrow s']$
be an expression (of type $\tau$) that wraps the input in a provided type. Then, for any expression $e_c$
(user code) such that $\emptyset; y:\tau \vdash e_c:\tau'$ and that does not contain any structural values
as sub-expressions, it is the case that $e_c[y\leftarrow e_s] \reduce^{*} v$ for some value $v$ and
also $\emptyset; \vdash v : \tau$.
\end{theorem}
\begin{proof}
We discuss the two parts of the proof separately as type prservation (Lemma~\ref{thm:rs-preservation})
and progress (Lemma~\ref{thm:rs-progress}).
\end{proof}

@@

@@

\section{Implementation}
\label{sec:impl}
~

\subsection{Parsing}
\label{sec:impl-parsing}
Turning JSON, CSV or XML to the idealized structured value presented here (esp. XML)

JSON - infering numbers from values (not just from actual value type - world bank example)

\subsection{Unions}
\label{sec:impl-unions}
Mainly because type providers can only provide objects

Unions become records with fields - so hiding null makes sense
also it means the open-world semantics

Heterogeneous collections

\subsection{Practical experience}
Error handling - solve issues just add another example

\section{Related and future work}

>>>>

\subsection{Infering types from values}
\label{sec:formal-inferval}

@@

Then for all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, let $e_s=e[x\leftarrow s']$
be an expression (of type $\tau$) that wraps the input in a provided type. Then, for any expression $e_c$
(user code) such that $\emptyset; y:\tau \vdash e_c:\tau'$ and that does not contain any structural values
as sub-expressions, it is the case that $e_c[y\leftarrow e_s] \reduce^{*} v$ for some value $v$ and
also $\emptyset; \vdash v : \tau$.
\end{theorem}
\begin{proof}
We discuss the two parts of the proof separately as type preservation (Lemma~\ref{thm:rs-preservation})
and progress (Lemma~\ref{thm:rs-progress}).
\end{proof}

@@

% ==================================================================================================
%
%   ###                                                                                     
%    #  #    # #####  #      ###### #    # ###### #    # #####   ##   ##### #  ####  #    # 
%    #  ##  ## #    # #      #      ##  ## #      ##   #   #    #  #    #   # #    # ##   # 
%    #  # ## # #    # #      #####  # ## # #####  # #  #   #   #    #   #   # #    # # #  # 
%    #  #    # #####  #      #      #    # #      #  # #   #   ######   #   # #    # #  # # 
%    #  #    # #      #      #      #    # #      #   ##   #   #    #   #   # #    # #   ## 
%   ### #    # #      ###### ###### #    # ###### #    #   #   #    #   #   #  ####  #    # 
%
% ==================================================================================================

\section{Implementation}
\label{sec:impl}

The theoretical model in Section~\ref{sec:formal} presents the core ideas behind the type 
providers for structured data formats. However, an important part of the success of the F\# Data
library is also its pragmatic approach, which makes it suitable for use with real-world data.

The formal model discusses some of the pragmatic choices, such as the preference for records over 
unions and the ability to treat $0$ and $1$ as both Booleans and numbers. In this section, we briefly
discuss some of the remaining practical concerns, starting with the handling of collections.

\subsection{Parsing strucutred values}
\label{sec:impl-parsing}

In this paper, we treat the XML, JSON and CSV formats uniformly as \emph{structural values}.
The definition of structural values is, indeed, rich enough to capture all three formats. The
structure is similar to JSON (it has primitive values, records and collections). We also added
\emph{named} records, which are needed for XML.
%
\begin{itemize}
\item When reading JSON, we directly create a corresponding structural value (with all records unnamed).
  Optionally, we also read numerical values (and dates) stored in strings.
\item When reading CSV, we read each row as an unnamed record and return them in a collection.
  Optionally, the read values are parsed as numbers or Booleans and missing values become \kvd{null}.
\item When reading XML, we create a named record for each node. Attributes become record fields and
  body becomes a special field named $\bullet$
\end{itemize}
%
The reading of XML documents is perhaps the most interesting. To demonstrate how the body is treated,
consider the following:
%
{\small{
\begin{verbatim}
  <root id="1">
    <item>Hello!</item>
  </root>    
\end{verbatim}
}}
%
\noindent
This XML becomes a record named \ident{root} with fields \ident{id} and $\bullet$. The nested element
contains only the $\bullet$ field containing the inner text:
%
\begin{equation*}
\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
\end{equation*}
%
When generating types for types inferred from XML, we also include a special case to remove the $\bullet$
node using the fact that this can be only a collection (of elements) or a primitive value.

Finally, the XML type provider in F\# Data includes an additional option to use \emph{global inference}.
In that case, the inference algorithm from Section~\ref{sec:formal-inferval} is replaced with an alternative
that unifies the types of all records with the same name. This is useful when the sample is, for example,
an XHTML document -- in this case, all occurrences of an element (\emph{e.g.}~{\small\ttfamily <a>})
are treated as the same type.

\begin{figure*}
\noindent
\begin{equation*}
\inference[(col)\;]
  { \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
    \sigma_i \tsep \sigma'_i \vdash \sigma''_i \quad
    \phi_i \tsep \phi'_i \vdash \phi''_i \qquad (\forall i\in \{ 1 .. k \}) }
  { \begin{array}{c}
    [\sigma_1,\psi_1 | \ldots | \sigma_k,\psi_k | \ldots | \sigma_n,\psi_n ] \tsep 
    [\sigma'_1,\psi'_1 | \ldots | \sigma'_k, \psi'_k | \ldots | \sigma'_m, \psi'_m ] \vdash\\[0.1em]
    [\sigma''_1, \psi''_1 | \ldots | \sigma''_k, \psi''_k | 
        \sigma_{k+1},\addopt{\psi_{k+1}} | \ldots | \sigma_{n},\addopt{\psi_{n}} |
        \sigma'_{k+1},\addopt{\psi'_{k+1}} | \ldots | \sigma'_{m}, \addopt{\psi'_{m}} ]
    \end{array} }
\quad
\begin{array}{rcl}
 \\
 \addopt{\phi} = \phi'&\narrow{\textnormal{such that}}&\phi \tsep 1? \vdash \phi' \\[0.1em]
 \phi \tsep \phi' \vdash \phi&\narrow{\textnormal{when}}&\phi :> \phi' \\[0.1em]
 \phi \tsep \phi' \vdash \phi'&\narrow{\textnormal{when}}&\phi' :> \phi
\end{array}    
\end{equation*}
\caption{Inference judgement that defines the common supertype of two heterogeneous collections}
\label{fig:subtyping-hetcol}
\end{figure*}

When introducing the XML type provider (Section~\ref{sec:providers-xml}), we briefly mentioned 
that F\# Data implements a special handling of heterogeneous collections. In the example, the provider
generated a type with \ident{Title} member of type \ident{string} (corresponding to a nested element
{\ttfamily\small <title>} that appears exactly once) and \ident{Items} member (returning a list of
values obtained from multiple {\ttfamily\small <item>} elements).

To capture this behaviour, we need to extend our earlier treatment of collections. Rather than storing
a single type for the elements as in $[\sigma]$, we store multiple possible element types. However,
this is not just a union type as we also store \emph{inferred multiplicity} of elements for 
each of the types:
%
\begin{equation*}
\begin{array}{rcl}
 \psi &\narrow{=}& 1 \lsep 1? \lsep \ast \\
 \sigma &\narrow{=}& ~\ldots \lsep [\sigma_1, \psi_1 | \ldots | \sigma_n, \psi_n ]
\end{array}
\end{equation*}
%
The multiplicities $1, 1?$ and $\ast$ represent \emph{exactly one}, \emph{zero or one} and \emph{zero or more},
respectively. Thus, the inferred type of the collection in the RSS feed would be
$[\ident{title}~\{ \ldots \}\;,1|\;\ident{item}~\{ \ldots \},\ast]$, which reads as a collection
containing exactly one \ident{title} record and any number of \ident{item} records.

The subtyping relation on heterogeneous collections specifies that a subtype can have fewer cases 
provided that they do not have to appear exactly once. A subtype can also have a stricter specification
of multiplicity (the ordering is $\ast :> 1? :> 1$).

Finding a common supertype of heterogeneous collections is analogous to the handling of union types.
The key rule is (\emph{col}) in Figure~\ref{fig:subtyping-hetcol}. It merges cases with the same tag
(by merging both the type and the multiplicity). For cases that appear only in one collection, we ensure
that the multiplicity is \emph{zero or one} or \emph{zero or more} using the auxiliary definition
$\addopt{-}$.

\subsection{Inferring and generating unions}
\label{sec:impl-unions}

The F\# Data type providers prefer types that do not contain unions (Theorem~\ref{thm:no-unions}).
The motivation for this is twofold. First, the current support for type providers in the F\# compiler
does not allow type providers to generate F\#-specific types (including discriminated unions). This
means that the provided type cannot use idiomatic F\# representation. Second, when accessing data,
using records aids the explorability -- when a value is a record, users can type ``.'' and most modern
F\# editors provide auto-completion list with members.

This choice has an important desirable consequence -- it allows treating a union with additional 
cases as a subtype of another union and, hence, the provided code is guaranteed to work on \emph{more}
inputs. Consider a type that is either a \ident{Person} record or just a string. 
The type provider exposes it as the following F\# class:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{PersonOrString}~=  \\[0.1em]
 \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl \ident{Person}\rangl \\[0.1em]
 \quad \kvd{member}~\ident{String}~:~\ident{option}\langl \ident{string}\rangl \\[0.1em]
\end{array}
\end{equation*}
%
If we provided an algebraic data type (discriminated union), the consumer would have to use pattern
matching that would cover \emph{two} cases. The above type forces the user to also handle a third
case, when both properties return \ident{None}. This also means that union types can silently skip
over \kvd{null} values.

That said, exposing a discriminated union (perhaps together with better tooling) is certainly an
attractive alternative that we plan to consider in the future. This can also be combined with type 
inference that chooses between records and union types based on some statistical metric (see
related work in Section~\ref{sec:related}).

\subsection{Practical experiences}

Sample is captured at compile time. It can come from the web or not (cache locally if you want
but from the web has nice implications).

We dont know what to do about schema change at runtime
(for some data sources, we could do more clever things though)

In practice..
Error handling - solve issues just add another example

@@

\section{Related and future work}
\label{sec:related}
~

Statistical metric makes the safety even messier (probabilistic!)
